---
title: "Covid Trends"
author: "Jose Tamez"
date: "4/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE, message = FALSE,comment = "#>")
library(stringr)

logisticcdf <- function(days,ro,to) 
{ 
  cdf <- 1.0/(1.0+exp(ro*(days-to))); 
  return(cdf);
}

logisticpdf <- function(days,ro,to) 
{ 
  pdf <- ro*exp(ro*(days-to))/((1.0+exp(ro*(days-to)))^2); 
  return(pdf);
}

```

## Forcasting COVID-19 Deaths

A simple model for the estimation of the maximum level of deaths due to COVID-19.

NOTE: I'm not an epidiomiologist so don't trust my anaysis. 

It is just a prediction based on: 

70% of SARS-CoV-2 exposure  
10% Infection efficacy ie. 10% of the exposure subjects turns into a Covid-19 case  
1% of Mortality  

For each country we will use the urban population  

I'll estimate the death rates based on current trends fitted to the logistic function.  
The last 9 days will be used for the estimations.  


```{r parameters}
# The expetec %  of deaths in each country

expectedtotalFatalities = 0.7*0.1*0.01

# The number of observations used for the trends
daysWindow <- 9

```


### Loading the data



```{r}


train <- read.csv("~/GitHub/COVIDTrends/Kaggle/train.csv", stringsAsFactors=FALSE)
train$ProviceCountry <- paste(train$Country_Region,train$Province_State,sep=":")

trainCountry <- train[train$Province_State == "",]

covid19_datapopulation <- read.csv("~/GitHub/COVIDTrends/Kaggle/covid19_data - population.csv", stringsAsFactors=FALSE)
rownames(covid19_datapopulation) <- paste(covid19_datapopulation$Country,":",sep="")


totalPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Population_2020,",",""))
totalUrbanPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Urban_pop_pct,"\\%",""))/100.0
totalUrbanPoulation[is.na(totalUrbanPoulation)] <- 1.0;
totalUrbanPoulation <- totalPoulation*totalUrbanPoulation


covid19_datapopulation$ExpectedFatalities <- expectedtotalFatalities*(
                                                  totalUrbanPoulation + 
                                                    0.1*(totalPoulation-totalUrbanPoulation)
                                              )


trainCountry$PerFatalities <- trainCountry$Fatalities/covid19_datapopulation[trainCountry$Country_Region,"ExpectedFatalities"]

trainCountry <- trainCountry[trainCountry$Fatalities > 0,]

expfatalities <- covid19_datapopulation$ExpectedFatalities/1.0e6
names(expfatalities) <- covid19_datapopulation$Country
expfatalities <- expfatalities[order(-expfatalities)]
plot.new()
op <- par(no.readonly = TRUE)
par(mar=c(8,4,4,4),pty="m")
#barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,horiz = TRUE,main="Expected Fatalities",xlab="MIllions")
barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,main="Expected Fatalities",ylab="Millions")
par(op)

```

Ploting some trends

```{r}
Country_Region <- names(table(trainCountry$Country_Region))
totaldeaths <- numeric()
for (ctr in Country_Region)
{
   totaldeaths <- append(totaldeaths,max(trainCountry[trainCountry$Country_Region == ctr,"Fatalities"]))
}
names(totaldeaths) <- Country_Region
totaldeaths <- totaldeaths[order(-totaldeaths)]

plot(trainCountry[trainCountry$Country_Region == names(totaldeaths[1]),"Fatalities"],main="Fatalities",xlab="Days",ylab="Fatalities")
for (ctr in names(totaldeaths[1:25]))
{
  linp <- trainCountry[trainCountry$Country_Region == ctr,"Fatalities"] 
  lines(linp)
  text(length(linp)-1,linp[length(linp)],ctr)
}

plot(trainCountry[trainCountry$Country_Region == names(totaldeaths[1]),"PerFatalities"],main="% Fatalities",xlab="Days",ylab=" % Fatalities")
for (ctr in names(totaldeaths[1:25]))
{
  linp <- trainCountry[trainCountry$Country_Region == ctr,"PerFatalities"]
  lines(linp)
  text(length(linp)-1,linp[length(linp)],ctr)
}



```


```{r results = "asis", dpi=600, fig.height= 5.0, fig.width= 6.0}

par(op)
cn = 1
par(mfrow=c(3,3),cex=0.5)
for (cn in c(1:17,19:30))
{
  endDay <- 14
  countryNumber <- cn
  
  mainName <- paste("March 24",names(totaldeaths[countryNumber]))
  
  datsetone <- trainCountry[trainCountry$Country_Region == names(totaldeaths[countryNumber]),"PerFatalities"]
  datsetchange <- c(datsetone[1],datsetone[2:length(datsetone)]-datsetone[1:(length(datsetone)-1)])
  lastobs <- length(datsetone)
  datsetchange <- 0.35*datsetchange + 
                  0.65*c(datsetone[1],0.5*(datsetone[3:lastobs]-datsetone[1:(length(datsetone)-2)]),datsetchange[lastobs])
  
  datasetone <- as.data.frame(cbind(days=c(1:length(datsetone)),fatalities = datsetone,newfatalities = datsetchange))
  numdays <- nrow(datasetone)

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  
  datasetone <- rbind(datasetone,cbind(days=c(180:185),fatalities = rep(1,6),newfatalities = rep(0,6)))
  
  daysrange <- append(c(startday:ndays),c((nrow(datasetone)-5):nrow(datasetone)))
  
  
  roestimate <- try(nls(fatalities ~ logisticcdf(days, ro, to),data = datasetone[daysrange,],control=list(warnOnly=TRUE)))

#  roestimate <- try(nls(fatalities ~ SSlogis(days, Asym, xmid, scal),data = datasetone[daysrange,],control=list(warnOnly=TRUE)))
  
  smo <- summary(roestimate)
  predictedTotalCases <- SSlogis(c(1:120),smo$coefficients[1,1],smo$coefficients[2,1],smo$coefficients[3,1])
  ymax <- max(c(predictedTotalCases,datasetone$fatalities))

  plot(predictedTotalCases,ylim=c(0,ymax),type="l",lty=2,xlab="days",ylab="% Fatalities",main=mainName)
  lines(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],lwd=3)
  
  newcases <- growlogistic(c(1:120),1.0/smo$coefficients[1,1],1.0/smo$coefficients[3,1],smo$coefficients[2,1])
  lines(10*newcases,lty=5)
  
  lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=4,lwd=2)
  

  endDay <- 7
  mainName <- paste("March 31",names(totaldeaths[countryNumber]))

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  
  daysrange <- append(c(startday:ndays),c((nrow(datasetone)-5):nrow(datasetone)))
  roestimate <- try(nls(fatalities ~ SSlogis(days, Asym, xmid, scal),data = datasetone[daysrange,],control=list(warnOnly=TRUE)))
  
  if (!inherits(roestimate, "try-error"))
  {
  
    smo <- summary(roestimate)
    predictedTotalCases <- SSlogis(c(1:120),smo$coefficients[1,1],smo$coefficients[2,1],smo$coefficients[3,1])
    ymax <- max(c(predictedTotalCases,datasetone$fatalities))
    plot(predictedTotalCases,ylim=c(0,ymax),type="l",lty=2,xlab="days",ylab="% Fatalities",main=mainName)
    lines(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],lwd=3)
    
    newcases <- growlogistic(c(1:120),1.0/smo$coefficients[1,1],1.0/smo$coefficients[3,1],smo$coefficients[2,1])
    lines(10*newcases,lty=5)
    
    lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=4,lwd=2)
  }
  else
  {
    plot(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],type="l",lty=2,xlab="days",ylab="% Fatalities",main=mainName)
    lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=4,lwd=2)
  }

  endDay <- 0
  mainName <- paste("April 6",names(totaldeaths[countryNumber]))

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  
  daysrange <- append(c(startday:ndays),c((nrow(datasetone)-5):nrow(datasetone)))
  roestimate <- try(nls(fatalities ~ SSlogis(days, Asym, xmid, scal),data = datasetone[daysrange,],control=list(warnOnly=TRUE)))
  
  smo <- summary(roestimate)
  predictedTotalCases <- SSlogis(c(1:120),smo$coefficients[1,1],smo$coefficients[2,1],smo$coefficients[3,1])
  ymax <- max(c(predictedTotalCases,datasetone$fatalities))
  plot(predictedTotalCases,ylim=c(0,ymax),type="l",lty=2,xlab="days",ylab="% Fatalities",main=mainName)
  lines(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],lwd=3)
  
  newcases <- growlogistic(c(1:120),1.0/smo$coefficients[1,1],1.0/smo$coefficients[3,1],smo$coefficients[2,1])
  lines(10*newcases,lty=5)
  
  lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=4,lwd=2)
    
}

```

