---
title: "Covid Forecasting"
author: "Jose Tamez"
date: "4/9/2020"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, error = FALSE, warning = FALSE, message = FALSE,comment = "#>")
library(stringr)


```


## Forcasting COVID-19 Deaths

A simple model for the estimation of the maximum percentage of deaths due to COVID-19.

NOTE: I'm not an epidiomiologist. 

The prediction based on a simple logistic model and:

70% of SARS-CoV-2 exposure  
10% Infection efficacy ie. 10% of the exposure subjects turns into a Covid-19 case  
1% of Mortality  

For each country we will use the urban population  

I'll estimate the death rates based on current trends fitted to the logistic function.  
The last 14 days will be used for the peak estimations.
Peak estimations will be done using:  

If the data has not reached the peak the plots will show estimations based on:  

1) The last reported 14 observations  
2) 14 observations from the last week  
3) 14 observations ending two weeks ago   
4) 14 observations ending three weeks ago  
5) 14 observations ending four weeks ago  


If the data reached the peak the plots will show:  

1) all the points after one week before the peak   
2) 14 observations before one week prior the peak   
3) 14 observations ending two weeks prior the peak    
4) 14 observations ending three week prior the peak  
5) 14 observations ending four week prior the peak    



We are also providing optimistic models. We assume that the # of fatalities is one third the expected.



Notes:  

1) If the current observations does not mach the estimated CDF, do not trust the predictions.  
2) If predicted CDF does not reach 1.0, then there is the potential for more infection in the future or COVID-19 is not as deadly as I estimated  


### Source code:  

https://github.com/joseTamezPena/COVID_Forecasting


```{r parameters}
# The expetec %  of deaths in each country

expectedtotalFatalities = 0.7*0.1*0.01
optGain <- 3
# The number of observations used for the trends
daysWindow <- 14
currentdate <- "April 10:" 

```


### Loading the data

The data is the time_covid19ing set from CSSE at Johns Hopkins University:

https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv



```{r , echo = FALSE}

time_covid19 <- read.csv("~/GitHub/COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

time_covid19[time_covid19$Country.Region=="Korea, South","Country.Region"] <- "South Korea"
time_covid19[time_covid19$Country.Region=="Czechia","Country.Region"] <- "Czech Republic (Czechia)"
time_covid19[time_covid19$Country.Region=="US","Country.Region"] <- "United States"

time_covid19Country <- time_covid19[time_covid19$Province.State == "",]
time_covid19Country$Province.State <- NULL
time_covid19Country$Lat <- NULL
time_covid19Country$Long <- NULL
rownames(time_covid19Country) <- time_covid19Country$Country.Region
time_covid19Country$Country.Region <- NULL

covid19_datapopulation <- read.csv("./Kaggle/covid19_data - population.csv", stringsAsFactors=FALSE)
rownames(covid19_datapopulation) <- paste(covid19_datapopulation$Country,":",sep="")


totalPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Population_2020,",",""))
totalUrbanPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Urban_pop_pct,"\\%",""))/100.0
totalUrbanPoulation[is.na(totalUrbanPoulation)] <- 1.0;
totalUrbanPoulation <- totalPoulation*totalUrbanPoulation


covid19_datapopulation$ExpectedFatalities <- expectedtotalFatalities*(
                                                  totalUrbanPoulation + 
                                                    0.1*(totalPoulation-totalUrbanPoulation)
                                              )

time_covid19CountryPerc <- time_covid19Country
time_covid19CountryPerc <- time_covid19Country/covid19_datapopulation[rownames(time_covid19Country),"ExpectedFatalities"]


expfatalities <- covid19_datapopulation$ExpectedFatalities/1.0e6
names(expfatalities) <- covid19_datapopulation$Country
expfatalities <- expfatalities[order(-expfatalities)]
plot.new()
op <- par(no.readonly = TRUE)
par(mar=c(8,4,4,4),pty="m")
#barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,horiz = TRUE,main="Expected Fatalities",xlab="MIllions")
barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,main="Expected Fatalities",ylab="Millions")
par(op)

```

Ploting some trends

```{r}
Country.Region <- rownames(time_covid19Country)
totaldeaths <- as.numeric(time_covid19Country[,ncol(time_covid19Country)])
names(totaldeaths) <- Country.Region
totaldeaths <- totaldeaths[order(-totaldeaths)]


ydata <- as.numeric(time_covid19Country[names(totaldeaths[1]),])
ydata <- ydata[ydata > 1e-6]
plot(ydata,main="% Fatalities",xlab="Days",ylab="Fatalities",xlim=c(1,ncol(time_covid19Country)))
text(length(ydata)-1,ydata[length(ydata)],names(totaldeaths[1]))

for (ctr in names(totaldeaths[1:30]))
{
  ydata <- as.numeric(time_covid19Country[ctr,])
  ydata <- ydata[ydata > 1e-6]
  lines(ydata)
  text(length(ydata)-1,ydata[length(ydata)],ctr)
}

totaldeaths  <- totaldeaths[!is.na(totaldeaths)]
totaldeaths  <- totaldeaths[totaldeaths > 100]

```

## Trends Modeling Functions

```{r, echo = FALSE}
source('./logfunctions.r')

```



## Estimating the peaks

```{r, echo = FALSE, results = "hide", dpi=300, fig.height= 5.0, fig.width= 7.0}

par(op)
cn = 2
par(mfrow=c(1,1),cex=0.75,mar=c(4,4,4,5))
for (cn in c(1:length(totaldeaths)))
{
  endDay <- 0
  countryNumber <- cn
  mainName <- paste(currentdate,names(totaldeaths[countryNumber]))
  
  datsetone <- as.numeric(time_covid19CountryPerc[names(totaldeaths[countryNumber]),])
  datsetone <- datsetone[datsetone > 0.0]
  datsetchange <- c(datsetone[1],datsetone[2:length(datsetone)]-datsetone[1:(length(datsetone)-1)])
  lastobs <- length(datsetone)
  datsetchange <- 0.35*datsetchange + 
                  0.65*c(datsetone[1],0.5*(datsetone[3:lastobs]-datsetone[1:(length(datsetone)-2)]),datsetchange[lastobs])
  
  datasetone <- as.data.frame(cbind(days=c(1:length(datsetone)),fatalities = datsetone,newfatalities = datsetchange))

  lastobs <- nrow(datasetone);
  numdays <- min(nrow(datasetone),which.max(datasetone$newfatalities)+7)

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  daysrange <- c(startday:lastobs)
  
  cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays)
  daystopeak <- cdffit$to-lastobs
  if (daystopeak <= 1)
  {
    cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays,10)
  }

  
  
  datasetOptimist <- datasetone;
  maxgain <- min(0.9/max(datasetOptimist$fatalities),optGain);
  

  if (!inherits(cdffit, "try-error"))
  {
    predictedTotalCases <- logisticcdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;
    ymax <- max(1,max(predictedTotalCases))

    plot(predictedTotalCases,ylim=c(0,ymax),type="l",lty=2,xlab="Days from first fatality",ylab="Fraction of the Total Expected Fatalities",main=mainName)
    lines(datasetone$days[c(1:lastobs)],datasetone$fatalities[c(1:lastobs)],lwd=5)
    lines(datasetone$days[c(1:lastobs)],10*datasetone$newfatalities[c(1:lastobs)],lty=3,lwd=4,col=2)

    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust
    lines(10*predictedTotalNewCases,lty=4,col="red")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit$to-lastobs;
    text(dmax + 10,0.95,
         paste("Days to peak:", sprintf("%3.0f",daystopeak)),
         cex=0.8)
    abline(v=dmax,col="red",lty=2)
    
    expfat <- covid19_datapopulation[names(totaldeaths[countryNumber]),"ExpectedFatalities"]*max(predictedTotalCases)
    text(100,0.5,
         paste("Total Expected Deaths:", sprintf("%3.0f",expfat)),
         cex=0.8)
    
    expfat <- covid19_datapopulation[names(totaldeaths[countryNumber]),"ExpectedFatalities"]*max(predictedTotalNewCases)
    text(100,0.45,
         paste("Maximum deaths per day:", sprintf("%3.0f",expfat)),
         cex=0.8)
  if (numdays ==  lastobs)
  {
    datasetOptimist <- datasetone;
    maxgain <- min(0.9/max(datasetOptimist$fatalities),optGain);
    datasetOptimist$fatalities <- maxgain*datasetOptimist$fatalities;
    datasetOptimist$newfatalities <- maxgain*datasetOptimist$newfatalities;
    cdffit2 <- logisitcfit(datasetOptimist[daysrange,],-0.1,ndays,10)  
    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit2$ro,cdffit2$to)/cdffit2$adjust

    predictedTotalCases <- logisticcdf(c(1:120),cdffit2$ro,cdffit2$to)/cdffit2$adjust;
    expfat <- covid19_datapopulation[names(totaldeaths[countryNumber]),"ExpectedFatalities"]*max(predictedTotalCases)/maxgain
    text(100,0.4,
         paste("Optimistic Deaths:", sprintf("%3.0f",expfat)),
         cex=0.8)
    
    
    lines(10*predictedTotalNewCases/maxgain,lty=5,lwd=2,col="pink")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit2$to-lastobs;
    text(dmax - 12,
         0.95,
         paste("Optimistic peak:",sprintf("%3.0f",daystopeak)),
         cex=0.8,
         col="pink")
    abline(v=dmax,col="pink")
  }

    for (endDay in c(7,14,21,28))
    {
      numdays <- which.max(datasetone$newfatalities);
      ndays <- numdays - endDay
      if (ndays > 7)
      {
        startday <- max(1,ndays - daysWindow)
        daysrange <- c(startday:ndays)
        cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays,10)  
        if (!inherits(cdffit, "try-error"))
        {
          newcases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;
          lines(10*newcases,lty=6,col=(endDay+14)/7,lwd= 1)
        }
      }
    }
  }
  else
  {
    plot(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],
         ylim=c(0,1.0),
         type="l",
         lty=2,
         xlab="days",
         ylab="% Fatalities",
         main=mainName)
  }
  legend("topright",
         legend = c("Estimated","Observed","New Fatalities","Estimated New","Optimistic New","One Week","Two Weeks","Three Weeks","Four Weeks"),
         col = c(1,1,2,2,"pink",3,4,5,6),
         lty = c(2,1,3,4,5,6,6,6,6),
         lwd = c(1,5,4,1,2,1,1,1,1),cex=0.75)
  z <- c(0:10)/100;
  axis(4, at=10*z,labels=round(z,digits=2),
  col.axis="blue", las=2, cex.axis=0.7, tck=-.01)
  mtext("Fraction of New Fatalities", cex=0.5,side=4, line=3, cex.lab=0.3, col="black",las=3)

}

```

