---
title: "Covid-19 Peak Forecasting"
author: "Jose Tamez"
date: "1/5/2020"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, error = FALSE, warning = FALSE, message = FALSE,comment = "#>")
library(stringr)


```


## Forecasting COVID-19 Deaths

A simple model for the estimation of the maximum percentage of deaths due to COVID-19 


============================================================================   

NOTE:  

1) I am not an epidemiologist.  
2) The predictions are not official, nor backed up by any organization nor government.  
3) I am providing them to get an idea of when the COVID-19 peaks may occur.   
4) I will update the predictions after 10:30pm CT every day   
5) They will be based on CSSE at Johns Hopkins University data release. Hence, the data may reflect the day before.  


"Disclaimer: Content from this website is STRICTLY ONLY for educational and research purposes and may contain errors. The model and data are inaccurate to the complex, evolving, and heterogeneous realities of different countries. Predictions are uncertain by nature. Readers must take any predictions with caution. Over-optimism based on some predicted end dates is dangerous because it may loosen our disciplines and controls and cause the turnaround of the virus and infection, and must be avoided."

Statement copied from: https://ddi.sutd.edu.sg/

============================================================================   

The prediction based on a simple logistic model and:

70% of SARS-CoV-2 exposure  
10% Infection efficacy i.e. 10% of the exposure subjects turns into a Covid-19 case  
2.0% of Mortality  

For each country we will use the 100% urban populationand 30% of the rural population

I will estimate the death rates based on current trends fitted to the two-logistic model function.  
The last days will be used for the peak estimations.

I'll show the model peak and the observed peak


=====================================================================================  
1/5/2021 Update:  
Change in algorithms to handle second waves less cluttered the plots

1/1/2021 Update:  
Change in algorithms to handle second waves, and CI based on bootstrapping

12/10/2020 Update:  
I'm changing the probability of getting covid-19 from 10% to 15%

8/7/2020 Update:  
The modeling function now has two logistics One for uptrend and the second for down trending.
The two functions are blended for the final fitting.

6/17/2020 Update:  
Changed intial fit conditions and fit stop criteria.

5/24/2020 Update:  

I'm updating the second wave estimations
It seems that we have learnead something so I'll change from 4.0% to 2.0% the death rate.

5/23/2020 Update:  

I'm getting ready for the second wave. 
So I have to adjust the mortality rate: From 1.0% to 4.0% based on NYC data

5/8/2020 Update:  

Code Update. I simplify the code and corrected minor bugs.

5/7/2020 Update:  

I replaced lowess smoothing to Friedman's SuperSmoother

4/28/2020 Update:  

The deaths per day now are smoothed by a median filter

4/16/2020 Update:  

Now the plots include the 95% confidence intervals of the expected peak. The estimations are done for the original estimations and the optimistic ones.




Notes:  

1) If the current observations do not match the estimated CDF, do not trust the predictions.  
2) If predicted CDF does not reach 1.0, then there is the potential for more infection in the future or COVID-19 is not as deadly as I estimated  


### Source code:  

https://github.com/joseTamezPena/COVID_Forecasting


```{r parameters}
# The expetec %  of deaths in each country

expectedtotalFatalities = 0.70*0.15*0.02
optGain <- 3
# The number of observations used for the trends
daysWindow <- 17

today <- Sys.Date()

currentdate <- paste(as.character(today),":") 

```


### Loading the data

The data is the time_covid19ing set from CSSE at Johns Hopkins University:

https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv



```{r , echo = FALSE}

time_covid19 <- read.csv("~/GitHub/COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

time_covid19[time_covid19$Country.Region=="Korea, South","Country.Region"] <- "South Korea"
time_covid19[time_covid19$Country.Region=="Czechia","Country.Region"] <- "Czech Republic (Czechia)"
time_covid19[time_covid19$Country.Region=="US","Country.Region"] <- "United States"

time_covid19Country <- as.data.frame(time_covid19[1,])
insn <- 1
datanames <- colnames(time_covid19)[-c(1:4)]
innames <- time_covid19Country$Country.Region;
rn <- 2
for (rn in 2:nrow(time_covid19))
{
  
  inn <- innames %in% time_covid19[rn,"Country.Region"]
  if (sum(inn) > 0)
  {
    time_covid19Country[inn,datanames] <- time_covid19Country[inn,datanames] + time_covid19[rn,datanames]
  }
  else
  {
    time_covid19Country <- rbind(time_covid19Country,time_covid19[rn,])
    insn <- nrow(time_covid19Country);
    innames <- time_covid19Country$Country.Region;
    rownames(time_covid19Country) <- innames
  }
}

time_covid19Country$Lat <- NULL
time_covid19Country$Long <- NULL
time_covid19Country$Province.State <- NULL
time_covid19Country$Country.Region <- NULL

covid19_datapopulation <- read.csv("./Kaggle/covid19_data - population.csv", stringsAsFactors=FALSE)
rownames(covid19_datapopulation) <- paste(covid19_datapopulation$Country,":",sep="")


totalPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Population_2020,",",""))
names(totalPoulation) <- covid19_datapopulation$Country
totalUrbanPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Urban_pop_pct,"\\%",""))/100.0
totalUrbanPoulation[is.na(totalUrbanPoulation)] <- 1.0;
totalUrbanPoulation <- totalPoulation*totalUrbanPoulation


covid19_datapopulation$ExpectedFatalities <- expectedtotalFatalities*(
                                                  totalUrbanPoulation + 
                                                    0.3*(totalPoulation-totalUrbanPoulation)
                                              )

time_covid19CountryPerc <- time_covid19Country
time_covid19CountryPerc <- time_covid19Country/covid19_datapopulation[rownames(time_covid19Country),"ExpectedFatalities"]


expfatalities <- covid19_datapopulation$ExpectedFatalities/1.0e6
names(expfatalities) <- covid19_datapopulation$Country
expfatalities <- expfatalities[order(-expfatalities)]
plot.new()
op <- par(no.readonly = TRUE)
par(mar=c(8,4,4,4),pty="m")
#barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,horiz = TRUE,main="Expected Fatalities",xlab="MIllions")
barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,main="Expected Fatalities",ylab="Millions")
par(op)

rownames(covid19_datapopulation) <- covid19_datapopulation$Country

```

Ploting some trends

```{r}
Country.Region <- rownames(time_covid19Country)
totaldeaths <- as.numeric(time_covid19Country[,ncol(time_covid19Country)])
names(totaldeaths) <- Country.Region
totaldeaths <- totaldeaths[order(-totaldeaths)]


ydata <- as.numeric(time_covid19Country[names(totaldeaths[1]),])
ydata <- ydata[ydata > 1e-6]
plot(ydata,main="# Fatalities",xlab="Days",ylab="Fatalities",xlim=c(1,ncol(time_covid19Country)))
text(length(ydata)-1,ydata[length(ydata)],names(totaldeaths[1]))

for (ctr in names(totaldeaths[1:30]))
{
  ydata <- as.numeric(time_covid19Country[ctr,])
  ydata <- ydata[ydata > 1e-6]
  lines(ydata)
  text(length(ydata)-1,ydata[length(ydata)],ctr)
}

totaldeaths  <- totaldeaths[!is.na(totaldeaths)]
totaldeaths  <- totaldeaths[totaldeaths > 5000]

```

## Trends Modeling Functions

```{r, echo = FALSE}
source('./logfunctions.r')


```



## Estimating the peaks

```{r, echo = FALSE, results = "hide", dpi=300, fig.height= 2.5, fig.width= 7.0}

totalDays = 450
names(totaldeaths)
par(op)
cn = which(names(totaldeaths) == "Italy")
#cn = 1 


par(mfrow=c(1,2),cex=0.35,mar=c(6,4,4,5))
estimated60 <- numeric(length(totaldeaths))
estimatedPeak <- estimated60
peakDates <- estimated60;
class(peakDates) <- "Date"
mainNames <- character(length(totaldeaths));

for (cn in c(1:length(totaldeaths)))
#for (cn in c(1:10))
{
  thepeakdates <- character(6);
  countryNumber <- cn
  mainName <- names(totaldeaths[countryNumber])
  mainNames[cn] <- mainName;
  
  datsetone <- as.numeric(time_covid19CountryPerc[names(totaldeaths[countryNumber]),])
  thedates <- colnames(time_covid19CountryPerc)[datsetone > 0.0]
  thedates <- str_replace(thedates,"X","")
  thedates <- str_replace_all(thedates,"\\.","/")
  thedates <- as.Date(thedates,"%m/%d/%Y")
  thedates <- as.Date(str_replace(thedates,"0020-","2020-"))

  datsetone <- datsetone[datsetone > 0.0]
  
  
  lastobs <- length(datsetone)
  datsetchange <- c(datsetone[1],datsetone[2:lastobs]-datsetone[1:(lastobs-1)])
  datsetchange[datsetchange < 0] <- 0;

  datasetone <- as.data.frame(cbind(days=c(1:lastobs),fatalities = datsetone,newfatalities = datsetchange))
  filteredNewFatalities <-  runmed(datasetone$newfatalities,5)
  filteredNewFatalities <- supsmu(c(1:lastobs),filteredNewFatalities)$y

  

  
#  lastobs <- nrow(datasetone) - 1;
  peakday <- which.max(filteredNewFatalities);
  if (datsetone[lastobs] < 0.025)
  {
    peakday <- lastobs;
  }
  mindef <- which.min(filteredNewFatalities[peakday:(lastobs-daysWindow-7)])+peakday-1;
  mindef <- min(mindef-daysWindow,3*lastobs/4)
  
  cat(lastobs,":",peakday,":",filteredNewFatalities[peakday],"Min day:",mindef,filteredNewFatalities[mindef],"\n");
  

  numdays <- min(lastobs,peakday-2*daysWindow+7)
  if (numdays < (lastobs-7))
  {
#    numdays <- which.max(filteredNewFatalities)
    if (max(datasetone$fatalities) < 0.25)
    {
      gain <- 0.25/max(datasetone$fatalities);
      datasetone$fatalities <- gain*datasetone$fatalities;
      datasetone$newfatalities <- gain*datasetone$newfatalities;
      covid19_datapopulation[mainName,"ExpectedFatalities"] <- covid19_datapopulation[mainName,"ExpectedFatalities"]/gain;
    }
  }
  plot(datasetone$newfatalities,main=paste("New Deaths:",mainName),xlab="Days since deaths>0",ylab="Fraction of expected deaths")


#  stddate <- max(peakday-lastobs/2,1)
  stddate <- 1
  cdffitglobal <- logisitcfit(datasetone,-0.05,peakday,5,daysrange=c(stddate:lastobs))
  #  cdffitglobal <- logisitcfit(datasetone,-0.05,peakday,5,daysrange=c(1:numdays))
#  cdffitglobal <- logisitcfit(datasetone,-0.05,peakday,5)
  

  print(c(peakday,numdays,lastobs))
  lines(logisticpdf(c(1:length(datsetone)),cdffitglobal$ro,peakday,cdffitglobal$alpha,cdffitglobal$lro,cdffitglobal$lto)/cdffitglobal$defecitRatio,lty=2,col="blue")
  too <- cdffitglobal$to
  startday <- min(cdffitglobal$to+31,as.integer(2*lastobs/3))
  adjini = 1.25*cdffitglobal$defecitRatio;
  daysrange <- c(startday:lastobs)
  cat(lastobs,":",too,": Start day:",startday,"Last Day:",lastobs,"\n");

  cdffit <- logisitcfit(datasetone,
                        cdffitglobal$ro,
                        too,
                        3.0*adjini,
                        adjini = adjini,
                        daysrange=daysrange)
 lines(cdffit$filterpdf,lty=1,col="red")

  lines(logisticpdf(c(1:length(datsetone)),cdffit$ro,cdffit$to,cdffit$alpha,cdffit$lro,cdffit$lto)/cdffit$defecitRatio,lty=5,col="green")
  daystopeak <- cdffit$lto-lastobs
#  cdffitn <- cdffitglobal;
  cdffitn <- cdffit;
  legend("topleft",
         legend = c("New Fatalities","Smooth","Global Fit","Last Trend Fit","Optimistic Trend Fit"),
         col = c(1,"red","blue","green","pink"),
         lty = c(NA,1,2,3,4),
         lwd = c(1,1,1,1,1),
         pch = c(1,NA,NA,NA,NA),
         cex=0.85)
  
#  datasetOptimist <- datasetone;
  athedates <- thedates[c(1:length(thedates)) %% 7 == 1]
  
  if (!inherits(cdffit, "try-error"))
  {
    bootest <- bootstraplogisitcfit(datasetone,cdffit,n=25,daysrange=daysrange)
    meanto <- median(bootest$to)
    meanro <- median(bootest$ro)
    meandr <- median(bootest$defratios)
    meanlto <- median(bootest$lto)
    meanlro <- median(bootest$lro)
    meanalpha <- median(bootest$alpha)
    
    torange <- quantile(bootest$lto,probs = c(0.025,0.5,0.975))

    dayloc <- numeric()
    maxdayloc <- numeric()
    maxloc <- numeric()
    predictedTotalCases <- logisticcdf(c(1:totalDays),cdffit$ro,cdffit$to,cdffit$alpha,cdffit$lro,cdffit$lto)/cdffit$defecitRatio;
    ymax <- max(1,max(predictedTotalCases))
    names(predictedTotalCases) <- thedates[1:totalDays]

    plot(predictedTotalCases,
         ylim=c(0,ymax),
#         ylim=c(0,0.03),
         type="l",
         lty=2,
         ylab="Fraction of the Total Expected Fatalities",
         xaxt="none",
         xlab="",
         main=mainName)
    lines(datasetone$days[c(1:lastobs)],datasetone$fatalities[c(1:lastobs)],lwd=3)
    for (n in 1:25)
    {
      if ((bootest$to[n] > 17) && (bootest$lto[n] > bootest$to[n]))
      {
      lines(c(lastobs:totalDays),
            logisticcdf(c(lastobs:totalDays),
                        bootest$ro[n],bootest$to[n],
                        bootest$alpha[n],bootest$lro[n],
                        bootest$lto[n])/bootest$defratios[n],
            lwd=1,
            lty=8,col="gray")
      }
      
    }
    lines(c(lastobs:totalDays),predictedTotalCases[lastobs:totalDays],lty=2,lwd=3,col="light blue")
    predictedTotalNewCases <- logisticpdf(c(1:totalDays),meanro,meanto,meanalpha,meanlro,meanlto)/meandr
    lines(10*predictedTotalNewCases,lty=4,col="red")
    dmax <- which.max(predictedTotalNewCases)
    peakloc <- dmax
    daystopeak <- cdffit$lto-lastobs;
    daystopeakrange <- torange-lastobs;

    peakdate <- thedates[lastobs] + as.integer(daystopeak+0.5);
    text(dmax + 10,0.95,
         paste("Days to peak:", sprintf("%4.1f",daystopeakrange[2])),
         cex=0.8)
    text(dmax + 10,0.90,
         sprintf("(%4.1f to %4.1f)",daystopeakrange[1],daystopeakrange[3]),
         cex=0.8)
    abline(v=dmax,col="red",lty=2)
    abline(v=cdffit$lto,col="pink",lty=2)

        
    totexpfat <- covid19_datapopulation[names(totaldeaths[countryNumber]),"ExpectedFatalities"];
    expfat <- totexpfat*
      logisticcdf(lastobs+60,meanro,meanto,meanalpha,meanlro,meanlto)/meandr;
    estimated60[cn] <- expfat;
    if (cdffitn$lto < (lastobs-1))
    {
      expatPeak <- totexpfat*datsetone[as.integer(cdffitn$lto) + 2];
    }
    else
    {
      expatPeak <- totexpfat*
        logisticcdf(meanto,meanro,meanto,meanalpha,meanlro,meanlto)/meandr;
    }
    estimatedPeak[cn] <- expatPeak;
    peakDates[cn] <- thedates[1]+as.integer(cdffitn$lto+0.5);
    text(90,0.75,
         paste("Current Deaths:", sprintf("%3.0f",max(time_covid19Country[names(totaldeaths[countryNumber]),]))),
         cex=0.8)
    text(90,0.7,
         paste("Estimated Total Deaths in 60 days:", sprintf("%3.0f from %5.0f Total Expected",expfat,totexpfat)),
         cex=0.8)
    
}
  thepeakdates <- as.character(thepeakdates)
  thepeakdates[1] <- paste("Estimated New Fat.")
  thepeakdates[2] <- paste("Optimistic New Fat.")
  legend("topright",
         legend = c("Total Estimated","Total Observed","New Observed","Flattening Trend"),
         col = c(1,1,2,"gray","purple"),
         lty = c(2,1,3,3,NA),
         lwd = c(1,3,3,3,1),
         cex=0.75)
  z <- c(0:10)/100;
  axis(4, at=10*z,labels=round(z,digits=2),
  col.axis="blue", las=2, cex.axis=0.7, tck=-.01)
  mtext("New Fatalities as Fraction of Total Expected", cex=0.35,side=4, line=3, cex.lab=0.2, col="black",las=3)

}

```
