---
title: "Covid Forecasting"
author: "Jose Tamez"
date: "4/9/2020"
output: 
  html_document: 
    fig_caption: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, error = FALSE, warning = FALSE, message = FALSE,comment = "#>")
library(stringr)


```


## Forcasting COVID-19 Deaths

A simple model for the estimation of the maximum percentage of deaths due to COVID-19.

NOTE: I'm not an epidiomiologist. 

The prediction based on a simple logistic model and:

70% of SARS-CoV-2 exposure  
10% Infection efficacy ie. 10% of the exposure subjects turns into a Covid-19 case  
1% of Mortality  

For each country we will use the urban population  

I'll estimate the death rates based on current trends fitted to the logistic function.  
The last 14 days will be used for the peak estimations.
Peak estimations will be done using:  

1) The last reported 14 observations  
2) 14 observations from the last week  
3) 14 observations ending two weeks ago   
4) 14 observations ending three weeks ago  
5) 14 observations ending four weeks ago  

We are also providing optimistic models. We assume that the # of fatalities is one third the expected.

For Mexico we set the optimistic factor to 5. For every other country is set to 3.


```{r parameters}
# The expetec %  of deaths in each country

expectedtotalFatalities = 0.7*0.1*0.01
optGain <- 3
mexicoGain <- 5
# The number of observations used for the trends
daysWindow <- 14
currentdate <- "April 8:" 

```


### Loading the data

The data is the training set from: Kaggle: COVID19 Global Forecasting (Week 4)

https://www.kaggle.com/c/covid19-global-forecasting-week-4/data



```{r , echo = FALSE}

train <- read.csv("./Kaggle/train.csv", stringsAsFactors=FALSE)
train$ProviceCountry <- paste(train$Country_Region,train$Province_State,sep=":")

train[train$Country_Region=="Korea, South","Country_Region"] <- "South Korea"
trainCountry <- train[train$Province_State == "",]

covid19_datapopulation <- read.csv("./Kaggle/covid19_data - population.csv", stringsAsFactors=FALSE)
rownames(covid19_datapopulation) <- paste(covid19_datapopulation$Country,":",sep="")




totalPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Population_2020,",",""))
totalUrbanPoulation <- as.numeric(str_replace_all(covid19_datapopulation$Urban_pop_pct,"\\%",""))/100.0
totalUrbanPoulation[is.na(totalUrbanPoulation)] <- 1.0;
totalUrbanPoulation <- totalPoulation*totalUrbanPoulation


covid19_datapopulation$ExpectedFatalities <- expectedtotalFatalities*(
                                                  totalUrbanPoulation + 
                                                    0.1*(totalPoulation-totalUrbanPoulation)
                                              )


trainCountry$PerFatalities <- trainCountry$Fatalities/covid19_datapopulation[trainCountry$Country_Region,"ExpectedFatalities"]

trainCountry <- trainCountry[trainCountry$Fatalities > 0,]

expfatalities <- covid19_datapopulation$ExpectedFatalities/1.0e6
names(expfatalities) <- covid19_datapopulation$Country
expfatalities <- expfatalities[order(-expfatalities)]
plot.new()
op <- par(no.readonly = TRUE)
par(mar=c(8,4,4,4),pty="m")
#barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,horiz = TRUE,main="Expected Fatalities",xlab="MIllions")
barplot(expfatalities[1:30],las=2,cex.names =0.70,cex.axis = 0.60,main="Expected Fatalities",ylab="Millions")
par(op)

```

Ploting some trends

```{r}
Country_Region <- names(table(trainCountry$Country_Region))
totaldeaths <- numeric()
for (ctr in Country_Region)
{
   totaldeaths <- append(totaldeaths,max(trainCountry[trainCountry$Country_Region == ctr,"Fatalities"]))
}
names(totaldeaths) <- Country_Region
totaldeaths <- totaldeaths[order(-totaldeaths)]

plot(trainCountry[trainCountry$Country_Region == names(totaldeaths[1]),"PerFatalities"],main="% Fatalities",xlab="Days",ylab=" % Fatalities")
for (ctr in names(totaldeaths[1:25]))
{
  linp <- trainCountry[trainCountry$Country_Region == ctr,"PerFatalities"]
  lines(linp)
  text(length(linp)-1,linp[length(linp)],ctr)
}


totaldeaths  <- totaldeaths[totaldeaths > 100]

```

## Trends Modeling Functions

```{r, echo = FALSE}
source('./logfunctions.r')

```



## Estimating the peaks

```{r, echo = FALSE, results = "hide", dpi=300, fig.height= 5.0, fig.width= 7.0}

par(op)
cn = 1
par(mfrow=c(2,2),cex=0.45,mar=c(4,4,4,5))
for (cn in c(1:length(totaldeaths)))
{
  endDay <- 0
  countryNumber <- cn
  mainName <- paste(currentdate,names(totaldeaths[countryNumber]))
  
  datsetone <- trainCountry[trainCountry$Country_Region == names(totaldeaths[countryNumber]),"PerFatalities"]
  datsetchange <- c(datsetone[1],datsetone[2:length(datsetone)]-datsetone[1:(length(datsetone)-1)])
  lastobs <- length(datsetone)
  datsetchange <- 0.35*datsetchange + 
                  0.65*c(datsetone[1],0.5*(datsetone[3:lastobs]-datsetone[1:(length(datsetone)-2)]),datsetchange[lastobs])
  
  datasetone <- as.data.frame(cbind(days=c(1:length(datsetone)),fatalities = datsetone,newfatalities = datsetchange))


  numdays <- nrow(datasetone)

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  daysrange <- c(startday:ndays)
  
  cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays)  

  
  
  datasetOptimist <- datasetone;
  maxgain <- min(0.9/max(datasetOptimist$fatalities),optGain);
  
  #Just for Mexico
  
  if (names(totaldeaths[countryNumber]) == "Mexico")
  {
     maxgain <- mexicoGain
  }
  
  datasetOptimist$fatalities <- maxgain*datasetOptimist$fatalities;
  datasetOptimist$newfatalities <- maxgain*datasetOptimist$newfatalities;
  cdffit2 <- logisitcfit(datasetOptimist[daysrange,],-0.1,ndays)  


  if (!inherits(cdffit, "try-error"))
  {
    predictedTotalCases <- logisticcdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;

    plot(predictedTotalCases,ylim=c(0,1.0),type="l",lty=2,xlab="Days from first fatality",ylab="Fraction of the Total Expected Fatalities",main=mainName)
    lines(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],lwd=5)
    lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=3,lwd=4,col=2)

    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust
    lines(10*predictedTotalNewCases,lty=4,col="red")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit$to-numdays;
    text(dmax+7,10*max(predictedTotalNewCases)+0.05,paste("Days to Peak:",sprintf("%3.0f",daystopeak)),cex=0.8)
    
    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit2$ro,cdffit2$to)/cdffit2$adjust
    lines(10*predictedTotalNewCases/maxgain,lty=5,lwd=2,col="pink")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit2$to-numdays;
    text(dmax+7,10*max(predictedTotalNewCases)/maxgain+0.05,paste("Opt: Days to Peak:",sprintf("%3.0f",daystopeak)),cex=0.8)

    for (endDay in c(7,14,21,28))
    {
      ndays <- numdays - endDay
      if (ndays > 7)
      {
        startday <- max(1,ndays - daysWindow)
        daysrange <- c(startday:ndays)
        cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays)  
        if (!inherits(cdffit, "try-error"))
        {
          newcases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;
          lines(10*newcases,lty=6,col=(endDay+14)/7,lwd= 1)
        }
      }
    }
  }
  else
  {
    plot(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],
         ylim=c(0,1.0),
         type="l",
         lty=2,
         xlab="days",
         ylab="% Fatalities",
         main=mainName)
  }
  legend("topright",
         legend = c("Estimated","Observed","New Fatalities","Estimated New","Optimistic New","One Week Ago","Two Weeks Ago","Three Weeks Ago","Four Weeks Ago"),
         col = c(1,1,2,2,"pink",3,4,5,6),
         lty = c(2,1,3,4,5,6,6,6,6),
         lwd = c(1,5,4,1,2,1,1,1,1),cex=0.5)
  z <- c(0:10)/100;
  axis(4, at=10*z,labels=round(z,digits=2),
  col.axis="blue", las=2, cex.axis=0.7, tck=-.01)
  mtext("Fraction of New Fatalities", cex=0.5,side=4, line=3, cex.lab=0.3, col="black",las=3)

}

```

## Estimation by Province or State

```{r, echo = FALSE}
trainProvince <- train[train$Province_State != "",]
#locations_population <- read.csv("~/GitHub/COVIDTrends/Kaggle/locations_population.csv", stringsAsFactors=FALSE)

library(readxl)
locations_population <- read_excel("Kaggle/locations_population.xlsx")

totalStatePoulation <- as.numeric(locations_population$UrabPop)
names(totalStatePoulation) <- locations_population$Province.State
totalStatePoulation <- totalStatePoulation[locations_population$Province.State != ""]


trainProvince$PerFatalities <- (trainProvince$Fatalities/totalStatePoulation[trainProvince$Province_State])/expectedtotalFatalities

trainProvince <- trainProvince[trainProvince$Fatalities > 0,]


Province_State <- names(table(trainProvince$Province_State))
totaldeaths <- numeric()
for (ctr in Province_State)
{
   totaldeaths <- append(totaldeaths,max(trainProvince[trainProvince$Province_State == ctr,"Fatalities"]))
}
names(totaldeaths) <- Province_State
totaldeaths <- totaldeaths[order(-totaldeaths)]
totaldeaths <- totaldeaths[totaldeaths>100]

```


## Estimating the State/Province Peaks

```{r, echo = FALSE, results = "hide", dpi=300, fig.height= 5.0, fig.width= 7.0}

par(op)
cn = 2
par(mfrow=c(2,2),cex=0.45,mar=c(4,4,4,5))
for (cn in c(1:length(totaldeaths)))
{
  endDay <- 0
  stateNumber <- cn
  mainName <- paste(currentdate,names(totaldeaths[stateNumber]))
  
  datsetone <- trainProvince[trainProvince$Province_State == names(totaldeaths[stateNumber]),"PerFatalities"]
  datsetchange <- c(datsetone[1],datsetone[2:length(datsetone)]-datsetone[1:(length(datsetone)-1)])
  lastobs <- length(datsetone)
  datsetchange <- 0.35*datsetchange + 
                  0.65*c(datsetone[1],0.5*(datsetone[3:lastobs]-datsetone[1:(length(datsetone)-2)]),datsetchange[lastobs])
  
  datasetone <- as.data.frame(cbind(days=c(1:length(datsetone)),fatalities = datsetone,newfatalities = datsetchange))
  numdays <- min(nrow(datasetone),which.max(datasetone$newfatalities)+7)

  ndays <- max(c(numdays - endDay,7))
  startday <- max(1,ndays - daysWindow)
  daysrange <- c(startday:ndays)
  cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays)  

  datasetOptimist <- datasetone;
  maxgain <- min(0.9/max(datasetOptimist$fatalities),optGain);
  datasetOptimist$fatalities <- maxgain*datasetOptimist$fatalities;
  datasetOptimist$newfatalities <- maxgain*datasetOptimist$newfatalities;
  cdffit2 <- logisitcfit(datasetOptimist[daysrange,],-0.1,ndays)  

  if (!inherits(cdffit, "try-error"))
  {
    predictedTotalCases <- logisticcdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;

    plot(predictedTotalCases,ylim=c(0,1.0),type="l",lty=2,xlab="Days from first fatality",ylab="Fraction of the Total Expected Fatalities",main=mainName)
    lines(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],lwd=5)
    lines(datasetone$days[c(1:ndays)],10*datasetone$newfatalities[c(1:ndays)],lty=3,lwd=4,col=2)
    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust

    lines(10*predictedTotalNewCases,lty=4,col="red")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit$to-numdays;
    text(dmax+7,10*max(predictedTotalNewCases)+0.05,paste("Days to Peak:",sprintf("%3.0f",daystopeak)),cex=0.8)
    

    predictedTotalNewCases <- logisticpdf(c(1:120),cdffit2$ro,cdffit2$to)/cdffit2$adjust
    lines(10*predictedTotalNewCases/maxgain,lty=5,lwd=2,col="pink")
    dmax <- which.max(predictedTotalNewCases)
    daystopeak <- cdffit2$to-numdays;
    text(dmax+7,10*max(predictedTotalNewCases)/maxgain+0.05,paste("Opt: Days to Peak:",sprintf("%3.0f",daystopeak)),cex=0.8)

    for (endDay in c(7,14,21,28))
    {
      ndays <- numdays - endDay
      if (ndays > 7)
      {
        startday <- max(1,ndays - daysWindow)
        daysrange <- c(startday:ndays)
        cdffit <- logisitcfit(datasetone[daysrange,],-0.1,ndays)  
        if (!inherits(cdffit, "try-error"))
        {
          newcases <- logisticpdf(c(1:120),cdffit$ro,cdffit$to)/cdffit$adjust;
          lines(10*newcases,lty=6,col=(endDay+14)/7,lwd= 1)
        }
      }
    }
  }
  else
  {
    plot(datasetone$days[c(1:ndays)],datasetone$fatalities[c(1:ndays)],
         ylim=c(0,1.0),
         type="l",
         lty=2,
         xlab="days",
         ylab="% Fatalities",
         main=mainName)
  }
  legend("topright",
         legend = c("Estimated","Observed","New Fatalities","Estimated New","Optimistic New","One Week Ago","Two Weeks Ago","Three Weeks Ago","Four Weeks Ago"),
         col = c(1,1,2,2,"pink",3,4,5,6),
         lty = c(2,1,3,4,5,6,6,6,6),
         lwd = c(1,5,4,1,2,1,1,1,1),cex=0.5)
  z <- c(0:10)/100;
  axis(4, at=10*z,labels=round(z,digits=2),
  col.axis="blue", las=2, cex.axis=0.7, tck=-.01)
  mtext("Fraction of New Fatalities", cex=0.5,side=4, line=3, cex.lab=0.3, col="black",las=3)

}

```


